%%
%% Abstract
%%
\begin{abstract}
    Systems incorporating language model components face a verification problem:
    these components fabricate, and verification has cost. How should a system
    allocate its verification budget? 
    Systems engineers building with language model
    components need to know what each level of verification investment buys.
    This paper provides the formal framework and empirical cost surface for making that decision.
    
    We prove that text-only observation with
    bounded supervision is structurally insufficient to verify epistemic honesty,
    regardless of model scale or training procedure (the impossibility is
    architectural, not a capability gap). 
    To overcome this, we introduce epistemic observability and construct a tensor interface that
    escapes this impossibility by exporting internal computational signals, such as entropy traces and
    attention summaries, that the model cannot independently control under current training objectives.
    Under a length-controlled evaluation where text-channel features collapse
    , per-token entropy remains discriminative , and tensor-guided judges outperform
    text baselines at every verification budget level (+2.5--3.9 percentage points, averaged
    across four architectures). Moreover, we show that tensor-guided judges
    outperform text baselines with lower verification budgets.
    % The text channel is
    % stronger than expected: response length alone achieves AUC
    % (Area Under the ROC Curve; 1.0 = perfect, 0.5 = random)
    % 0.88--0.96, but is trivially gameable.
    % After controlling for length, the tensor residual adds AUC 0.60--0.70:
    % modest, but occupying a fundamentally different position in the adversarial
    % landscape. Behavioral signals can be adapted by training without altering the
    % computation; computational signals cannot be independently controlled
    % under current training objectives, the same structural pattern that makes
    % code plagiarism detection (MOSS) robust. Preliminary observations
    % identify a format-dependent cost surface: scaffolding ratio and rigidity
    % vary across content formats (natural text, code, mathematical proofs) and
    % predict which judge strategy is cost-effective per domain. The
    % contribution is the map: 
    \end{abstract}
