%% ==========================================================================
%% Conclusion
%% ==========================================================================
\section{Conclusion}\label{sec:conclusion}

Systems incorporating language model components must decide how much to spend
on verification and where to spend it. This paper provides the cost surface
for that decision.

We have shown that epistemic honesty is unverifiable for systems operating under
text-only observation models with bounded supervision. This is not a capability
limitation to be solved with scale, nor a training failure to be solved with better
RLHF. It is a structural property of the interface: the channel through which
these systems communicate discards the epistemic metadata that would make
verification possible. This impossibility establishes the floor of the cost
surface: below this line, text-channel investment cannot improve assurance.

The impossibility has a shape. Policies conditioning only on queries cannot
satisfy epistemic honesty across ambiguous world states (\autoref{thm:representational}). Even
grounded architectures cannot learn honest behavior when supervisors cannot
distinguish plausible fabrications from truth (\autoref{thm:learnability}). Stacking judges
does not escape the constraint (\autoref{cor:layered}). Verification cost grows
superlinearly with response complexity (\autoref{lem:superlinear}). Responsibility concentrates
with the system owner (\autoref{cor:responsibility}).

But impossibility results are maps, not walls. The Fischer-Lynch-Paterson
theorem did not end distributed systems; it clarified what assumptions had to
change. Our result plays an analogous role. Systems that export structured
epistemic state (provenance, entropy traces, attention topology) escape the
impossibility class. The tensor interface demonstrates this escape is
achievable with current architectures and minimal overhead.

The empirical cost surface is honest. Under bounded verification,
tensor-guided judges at 10\% budget outperform text-guided judges at 30\%
budget, which is a 3$\times$ reduction in verification cost. The text channel is
stronger than expected (response length alone: AUC 0.88--0.96) but trivially
gameable. The tensor residual (AUC 0.60--0.70) is modest but occupies a
fundamentally different position in the adversarial landscape: computational
signals cannot be independently controlled without adversarial optimization
targeting the signal itself. Gaming the metric requires changing the thing
being measured.

The engineering question is not whether to verify, but how much verification
to buy and at which tier. A brainstorming tool needs Tier~0. A medical system
needs Tier~2. The cost surface tells you what each investment buys, and
preliminary observations indicate that content format modulates the tradeoff.

The work ahead is not exhorting models to try harder at honesty. It is
building systems that make honesty verifiable by design, with verification
budgets allocated according to the cost surface rather than spread uniformly
across outputs. The tools exist: vector clocks, cryptographic signatures,
content-addressable storage. The architectural patterns exist. What remains is
applying them to the epistemic dimension with the same rigor we learned to
apply to causality and integrity in distributed systems.

The interface is the problem. The interface is where the solution lives.
