%% ==========================================================================
%% CUT CONTENT: Tensor-Gated Composition (Section 4.4)
%% Removed from SOSP submission for page budget.
%% Preserved for arXiv version and/or blog post.
%% ==========================================================================

\subsection{Tensor-Gated Composition}

The tensor interface generalizes to composed systems. Recent work on Recursive
Language Models (RLMs) demonstrates that LLM outputs can feed into subsequent
LLM calls, enabling unbounded context through recursive
decomposition~\cite{zhang2025recursivelanguagemodels}. However, RLM reports failure modes
characteristic of composition without epistemic state: models abandon correct
intermediate answers, make redundant verification calls, and lose track of what
they have already solved.

The tensor provides a missing primitive: epistemic metadata that can propagate
across composition boundaries. We demonstrate this with tensor-gated
composition, where entropy thresholds control whether outputs propagate to
subsequent calls.

\fakepara{Experimental setup.}
We test three query chains: two fabrication prompts (``What is Glavinsky
syndrome?'' and ``Describe the Brennan-Kowalski theorem'') and one grounded
control (``What is the capital of France?''). For each, we generate an initial
response with tensor extraction, then use an entropy threshold ($\tau = 0.6$)
to gate whether the followup query proceeds.

\fakepara{Results.}
The grounded query produces entropy 0.02, well below threshold; the followup
proceeds and generates accurate content about Parisian landmarks. Both
fabrication queries produce entropy above 0.8; the gate blocks followup queries
that would have requested treatments for the nonexistent syndrome or
implications of the fictional theorem.

This is a minimal demonstration, not a complete solution. But it establishes
that tensor metadata, when used to gate composition, can prevent fabrication
chains from propagating. The supervisor, whether human or automated, gains
information that the text-only interface discards.

\fakepara{Symmetric tensor interface.}
Full compositional integrity requires a symmetric interface: not just
$f(\text{prompt}) \to \text{Tensor}$, but
$f(\text{Tensor}_{\text{in}}) \to \text{Tensor}_{\text{out}}$, where each call
receives the prior call's epistemic context and extends it. For fresh prompts,
the input tensor is empty. For composed calls, the input tensor carries
accumulated provenance and epistemic state. This design enables epistemic
traces to thread through recursive pipelines, providing verification handles at
each composition boundary.
