\section{Limitations and Future Work}\label{sec:discussion}

We are explicit about what this paper does and does not claim. The cost surface
is characterized; the investment tiers are mapped; but several important
questions remain open.

\fakepara{What we claim.}
\begin{itemize}
\item The impossibility is structural, not a training failure: text-only interfaces
  with bounded supervision cannot verify epistemic honesty regardless of model
  capability. This establishes the floor of the cost surface.
\item The tensor interface provides a different investment tier: concrete
  computational signals exist in current architectures that distinguish
  grounded from ungrounded generation, and these signals cannot be independently
  gamed.
\item We do not claim that current tensor signals solve epistemic verification. We
  claim that they occupy a fundamentally different position in the
  cost-robustness tradeoff than text-channel signals.
\item Current architectures already carry the relevant information; they just
  discard it at the interface.
\end{itemize}

\fakepara{What we do NOT claim.}
\begin{itemize}
\item The tensor interface is unfakeable under adversarial training.
  A model trained to minimize the correlation between entropy and
  truthfulness might learn to present confident distributions even
  when fabricating. This is an open question.
\item Composition preserves epistemic properties. If a model's output becomes
  input to another model (as in recursive LLM-mediated reasoning), the tensor
  signals from the first model may not propagate meaningfully.
\item The tensor interface is the optimal design. We demonstrate existence, not
  optimality. Better interfaces may achieve stronger guarantees with lower
  overhead.
\item These findings generalize to all architectures. Our empirical work spans
  four open-weight model families (models whose internal parameters are publicly
  available); closed-weight (API-only) and significantly larger models
  may exhibit different signal profiles.
\end{itemize}

\fakepara{Future work: Adversarial robustness.}
Can training find weights that produce confident tensors for fabricated content?
This is the adversarial analog of our structural results. Recent work by
Winninger et al.~\cite{winninger2025usingmechanisticinterpretabilitycraft}
demonstrates that mechanistic interpretability tools can be used to identify
internal subspaces associated with safety behavior and craft attacks that target
those subspaces directly. This confirms that internal signals are rich enough to
be both attacked and defended, which is precisely why exporting them matters.
The intuition that ``you cannot fake the computation you actually performed''
may break down if the model is trained end-to-end to minimize the tensor signal
on fabrications. Preliminary analysis suggests that linguistic confidence and
tensor confidence are decoupled in natural model outputs, models that hedge
linguistically do not systematically show higher tensor entropy, but whether
adversarial training can force this coupling remains open.
Information-theoretic bounds on signal manipulation are needed.

\fakepara{Future work: Compositional integrity.}
Does tensor composition preserve epistemic properties in recursive settings?
When LLM outputs feed into subsequent LLM inputs, the epistemic trace must
either be preserved, regenerated, or discarded. Designing interfaces that
maintain epistemic integrity across composition is an open architectural problem.

\fakepara{Cross-architecture generalization.}
Our bounded verification experiment tests four model families (OLMo-3, Llama-3.1,
Qwen3, Mistral) and finds consistent tensor advantage across all of them. Mean
pairwise Spearman $\rho = 0.762$ for entropy signals on the same queries
indicates the signal is architectural rather than model-specific. Extending to
closed-weight models that expose logits through API (e.g., via log-probability
endpoints) is a natural next step.

\fakepara{Future work: User-facing rendering.}
How should tensor information be presented to end users? Raw entropy traces are
not interpretable; summarization into confidence bands, uncertainty flags, or
visual indicators requires HCI research beyond the scope of this paper.

\fakepara{Model specificity.} Our topological analysis (\autoref{sec:escape})
uses OLMo, which permits full inspection of intermediate layer representations. The bounded verification
experiment (\autoref{tab:budget}) tests four architectures spanning 4B--8B
parameters (billions of learned numerical weights) and four distinct training pipelines. The self-report inversion
finding is universal across all four. We do not claim exhaustive coverage of
all model families, but the consistency across architecturally diverse models
suggests the findings are not idiosyncratic.

\fakepara{Domain-specific calibration.} The tensor signal is not uniformly discriminative across domains. In preliminary experiments on citation generation, entropy was \emph{lower} for fabricated citations than for real ones where the model generates plausible fakes more fluently than it retrieves specific facts. This confirms that the tensor measures generation confidence (``epistemic truth''), not truthfulness directly.  However, this blind spot is precisely where bounded verification is cheapest: citation verification has low compositional complexity (check the DOI, query the database,) which places it well within a bounded supervisor's budget. The tensor and the bounded judge have complementary failure modes as the tensor catches uncertain fabrication where verification is expensive and the judge catches confident fabrication where verification is cheap.  Composing them allocates verification resources to the failure class each is suited for, rather than spreading a finite budget across the full space.

\fakepara{Format-constraint variation: preliminary observations.}
The cost surface is not flat across content formats.
Tokenization analysis of generated code reveals that BPE compression
concentrates syntactic scaffolding (keywords, operators, whitespace) into
11--19\% of output tokens, while semantic content (variable names, literals,
strings) occupies 61--72\%. However, semantic content in familiar code patterns
is itself highly predictable: conventional variable names (\texttt{left},
\texttt{right}, \texttt{mid} in binary search) are as low-entropy as keywords,
constituting a second scaffolding layer we term \emph{semantic scaffolding}.
When both layers are considered, the entropy trace for familiar code is
predominantly flat, with spikes concentrated at genuine decision points---where
the model could not pattern-match and had to choose. We have not yet measured
discrimination AUC across content formats; what we report is the structural
observation that scaffolding ratio varies dramatically across output types and
predicts where entropy-based triage should be most efficient. If the
relationship holds, tensor-guided verification may be \emph{more}
cost-effective for code than for natural text, because higher scaffolding ratios
concentrate the informative signal into fewer tokens.
We do not claim to have mapped the full manifold. We claim it exists, that its
parameterization is observable, and that mapping it is the natural next step
for systems engineers choosing verification strategies across content domains.

\fakepara{Sufficiency gap.} The three architectural principles (State
Exteriority, Verification Independence, Provenance Binding) are \emph{necessary}
conditions for escaping the impossibility. We do not prove they are
\emph{sufficient}. A system satisfying all three might still fail epistemic
honesty through implementation bugs, adversarial inputs, or unforeseen failure
modes. Sufficiency requires additional formal treatment.

\fakepara{The engineering decision.}
How much epistemic assurance a system needs is an engineering decision, not a
research question. A brainstorming tool needs none. A medical summary system
needs substantial assurance. A legal research system needs the highest
tier available. This paper provides the cost surface for making that decision:
what each level of verification investment buys, where the text-channel ceiling
lies, what the tensor interface adds, and how content format modulates the
tradeoff. The contribution is the map. The territory is the system you are
building.

