%% ==========================================================================
%% SECTION 1: INTRODUCTION
%% ==========================================================================
\balance
\section{Introduction}\label{sec:intro}

Large Language Models (LLMs) already play a significant role
in day-to-day life. Despite their increasing importance,
especially in critical daily functions, these models
``hallucinate'', \ie, they generate plausible yet factually incorrect
or nonsensical information. While the term ``hallucination''
implies pathology: a glitch in an otherwise functional system,
we argue this is an \emph{expected behavior of the system}
because these models are optimized for coherence. 

Language models optimized for coherence produce
coherent output. When the training data provides grounding, that
output tends to track truth. When it does not, the same optimization
produces fluent, confident fabrication.
Thus, there has been a lot of recent work in reducing hallucinations
while optimizing for coherence.
However, it is a 
formally proven fact that these hallucinations can not be completely
eliminated~\cite{xu2024hallucination}. 

% Anyone who has used a large language model for research knows
% the ritual: check every citation. Not spot-check. \emph{Every citation}.
% Because the model will, with complete confidence, invent authors,
% journal names, publication years, and DOIs for papers that do not
% exist.

% The field calls this ``hallucination,'' a term that implies
% pathology: a glitch in an otherwise functional system. We argue the
% framing is wrong. A language model optimized for coherence produces
% coherent output. When the training data provides grounding, that
% output tends to track truth. When it does not, the same optimization
% produces fluent, confident fabrication. This is not a bug. It is the
% \emph{expected behavior} of a system that was never designed to distinguish
% the two cases.

The key problem is that \emph{we cannot tell when these models ``hallucinate''}.
Consequently, we utilize a myriad of techniques to validate the generated output.
However, the key challenge with validating the generated output is that
it is not known a priori which parts, if any, of the generated output
are fabrications.

To discern fabrications, we need access to the internal computation of the LLMs.
Current architectures carry
internal signals, such as entropy traces, attention geometry, and token-level
log-probabilities, that distinguish grounded generation from
fabrication. 
However, the relevant telemetry of the internal state of the generation is not
exposed to the users. 

The fundamental limitation is the existing text-only output interface of LLMs.
The generated output is merely the testimony of the generation process: the model's chosen output.
It does not contain the relevant telemetry of the internal state of the generation.
As the text-only output interface is optimized for producing coherent output,
it does not provide \emph{epistemic honesty}, the property of a system
to faithfully reflect the system's internal representation that something 
is known, unknown, or uncertain. Thus, we can not solely rely on the text-only output
itself to self-report the confidence of correctness for the generated output.
We prove that for predictive token generation systems such as LLMs operating under
text-only observation with bounded supervision, epistemic honesty is
formally unverifiable. 

% But the text-only interface through which users and
% supervisors observe the system discards this state. What remains is
% testimony: the model's chosen output. What is lost is telemetry:
% measurements of the computation that produced it. The model
% controls its testimony but it does \emph{not} control its telemetry. Yet the
% interface exports only the former.

This creates an observability gap with direct consequences for
verification. A supervisor, human or automated, receiving only text must spend its entire
verification budget attempting to reconstruct, from the output alone,
what the model's own computation already knew. The budget is
consumed by reconstruction rather than concentrated on the cases
where verification is needed. 
% We formalize this gap and
% prove that it constitutes a structural barrier, not merely a practical
% inconvenience.

% Large language models are examples of predictive token generation systems:
% they generate outputs by sampling from a probability distribution over
% possible continuations of the input text.

% Epistemic honesty is the property that a system's outputs faithfully reflect
% the system's internal representation that something is known, unknown, or uncertain.
 
% It is impossible, in the same sense that the
% Fischer-Lynch-Paterson result proved consensus impossible under
% asynchrony with one faulty process~\cite{flp}.

% To be precise about the type of claim: this is a \emph{verification}
% impossibility. We do not prove that fabrication is inevitable:
% \citeauthor{xu2024hallucination}~\cite{xu2024hallucination} prove
% a related but distinct result along those lines. We prove that even a
% finite model with perfect internal knowledge of its own epistemic state
% cannot have that knowledge \emph{verified} through text-only interfaces
% with bounded supervision. The model may know it is fabricating.
% The supervisor cannot tell.

% The impossibility decomposes into two results. The first is
% intuitive: a system that sees only the question cannot know whether
% the question is answerable in the real world, and therefore cannot
% know when honesty requires answering versus abstaining. Formally, any
% policy conditioning solely on queries cannot satisfy epistemic honesty
% across ambiguous world states (\textbf{Representational
% Impossibility}, \autoref{thm:representational}). The second result addresses the natural
% escape attempt: give the model access to external information.
% Retrieval-augmented architectures can in principle represent
% world-state-dependent behavior, but even these cannot \emph{learn}
% honest policies via preference optimization when supervisors are
% bounded, because plausible fabrications are observationally
% indistinguishable from truthful responses (\textbf{Learnability
% Impossibility}, \autoref{thm:learnability}).

% We validate the premise empirically: the internal signals exist and
% are being discarded. Across four model architectures (OLMo-3 7B,
% Llama-3.1 8B, Qwen3 4B, Mistral 7B), internal tensor
% signals (per-token entropy, attention statistics, and
% log-probabilities) reliably discriminate grounded from ungrounded
% generation. The discrimination is architectural: it appears in every model family tested. Yet when asked
% to self-report confidence through text, every model tested claimed
% \emph{higher} confidence on fabrications than on knowable facts. The
% models carry the epistemic state needed for verification. The
% interface throws it away. The testimony lies. The telemetry does not.

% This is a familiar class of problem in systems design. Distributed
% systems faced an analogous challenge: processes carried internal
% causal state (logical clocks, dependency graphs, and happened-before
% relations) that was invisible to external observers. Without access to
% this state, consistency guarantees were impossible to verify from
% message contents alone. The epistemic state of a language model plays
% the same structural role as causal state in a distributed system: it
% is internal metadata that external observers require for verification,
% and that the interface historically discards. The systems community
% did not solve consistency by making processes ``try harder.'' It solved
% consistency by making causal state observable with the export of
% vector clocks, logical timestamps, and causal traces alongside messages.

To address this gap, we propose that the interface must expose
model's epistemic state required for verification and validation.
% We apply the same principle to the epistemic dimension. The
% impossibility result characterizes what must change; the following
% architectural principles follow directly from the theorems.
We introduce epistemic observability for LLMs that exposes epistemic
telemetry of the model produced during the generative procedure.
As part of epistemic observability, we introduce two key components, entropy traces
and attention summaries, that encode the epistemic metadata.
% We identify three properties required to escape the impossibility class:
% \textbf{State Exteriority} (grounding generation in external world
% state, addressing \autoref{thm:representational}), \textbf{Verification Independence}
% (verification signals orthogonal to the generation channel, addressing
% \autoref{thm:learnability}), and \textbf{Provenance Binding} (structural links from
% output claims to verifiable sources, addressing the superlinear
% verification cost of \autoref{lem:superlinear}). 
The key insight behind epistemic observability is that the model's
internal geometry contains architecture-independent internal signals
capable of distinguishing grounded from fabricated generation
and that exposing this information overcomes the limitation of the text-only interface.

We demonstrate a concrete
\emph{tensor interface} for epistemic observability that exports epistemic metadata (entropy
traces and attention summaries) alongside generated text, requiring
no architectural changes to current transformers.
We show that our tensor interface can achieve upto 10\% improvement
in end-to-end accuracy across four different architectures.% on 800 query--response pairs.
% On 800 query--response pairs across four
% architectures, tensor-guided verification at 10\% budget outperforms
% text-guided verification at 30\% budget (82.1\% vs.\ 80.4\%
% end-to-end accuracy). Composing tensor-guided and bounded-lookup
% judges achieves 92.5\% at 30\% budget, demonstrating that
% complementary failure modes enable principled budget allocation.
% Ground truth is validated against human review (93.8\% agreement).

This paper makes the following contributions:

\begin{itemize}
\item \textbf{Impossibility result of epistemic verification for text-only interface.}
  \autoref{sec:formal} presents two theorems that prove that
  epistemic honesty is unverifiable for predictive token generation
  systems, including LLMs, operating under text-only observation models with bounded
  supervision. The impossibility is structural: it holds regardless
  of model scale, training procedure, or prompt engineering, and
  cannot be escaped by stacking supervisors.

\item \textbf{Epistemic Observability.}
  We introduce the concept of epistemic observability, \ie, metrics
  and traces from the internal generation procedure of the models
  that carry epistemic metadata about the state of the model.

\item \textbf{A tensor interface for epistemic observability.} We
  demonstrate a minimal interface extension that exports epistemic
  telemetry alongside text with no architectural changes to 
  current transformer implementations.
\end{itemize}

% We proceed as follows. \autoref{sec:background} establishes background on how truth
% regularities emerge in training data and where current practice breaks
% that inheritance. \autoref{sec:empirical} presents empirical evidence that internal
% epistemic state exists and is discarded by the text-only interface.
% \autoref{sec:formal} formalizes the impossibility. \autoref{sec:escape} characterizes the
% architectural escape and demonstrates the tensor interface.
% \autoref{sec:related} discusses related work.