\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsfonts, mathtools}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{float}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{subcaption}
\usepackage{cite}
\usepackage{setspace}

% Code listing style
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    frameround=tttt,
    backgroundcolor=\color{gray!5},
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{red},
    columns=fullflexible,
    escapeinside={(*}{*)},
    literate={_}{\textunderscore}1,
    mathescape=true,
}

\title{Supplementary Material: \\ Epistemic Observability in Language Models \\
    \normalsize An Impossibility Result with Tensor Escape}

\author{Submitted to SOSP 2026}

\date{}

\begin{document}

\maketitle

\begin{abstract}
This supplementary material provides formal specifications, mechanized proofs, and detailed methodology supporting the main paper ``Epistemic Observability in Language Models.''

\noindent \textbf{Contents:}
\begin{enumerate}
    \item TLA+ formal specifications of the impossibility theorem and tensor escape
    \item Machine-checked Lean4 proofs of the core theorems
    \item Experiment 27/27b detailed methodology (stratified evaluation, calibration protocol)
    \item Topological data analysis (TDA) methodology and results
    \item Supporting analyses and failure mode catalogs
\end{enumerate}

\end{abstract}

\newpage
\tableofcontents
\newpage

% ===========================================================================
% Section 1: TLA+ Specifications
% ===========================================================================

\section{Formal Specifications (TLA+)}

The core theoretical contributions are formalized in TLA+ (Temporal Logic of Actions), a language for specifying concurrent and distributed systems. Two complementary specifications model the impossibility under text-only observation and the escape under tensor interfaces.

\textbf{Epistemic status:} These are \emph{conditional formal results}. TLC verifies properties \emph{within our formal model}, where the axioms that (a) fabrications have distinguishable provenance and (b) text-only judges cannot access provenance are encoded by construction. The specifications demonstrate that our definitions are internally consistent and that the impossibility/escape follow from the stated axioms. The empirical question---whether real tensor signals satisfy these axioms---is addressed in the main paper's evaluation section.

\textbf{Note:} The excerpts below illustrate the modeling approach. Complete, runnable specifications are available in the artifact repository at \texttt{tla/EpistemicImpossibility.tla} and \texttt{tla/epistemic\_tensor.tla}. Both specifications have been validated with TLC model checker and verified to be correct.

\subsection{Specification 1: Text-Only Impossibility}

The file \texttt{tla/EpistemicImpossibility.tla} models the text-only observation regime:

\begin{verbatim}
MODULE EpistemicImpossibility
EXTENDS Naturals, Sequences, TLC

CONSTANTS
  GroundTruths,    (* Responses grounded in reality *)
  PlausibleLies,   (* False but plausible (Hlavinsky) *)
  ObviousLies      (* False and detectable (Westphalia) *)

VARIABLES internal_state, vector_clock, interface_out

(* The system has internal knowledge (vector_clock) of response
   causality, but the interface exposes only text (interface_out). *)

Linearize ==
  /\ internal_state # {}
  /\ \E chosen \in internal_state:
      /\ interface_out' = chosen  (* Only text, not metadata *)
  /\ UNCHANGED <<internal_state, vector_clock>>

(* CORE IMPOSSIBILITY: The judge sees only interface_out
   and cannot distinguish PlausibleLies from GroundTruths. *)

JudgeVerify(text) ==
  text \notin ObviousLies  (* Baseline heuristics only *)

Indistinguishable ==
  \E h \in PlausibleLies:
    /\ JudgeVerify(h) = TRUE           (* Judge accepts *)
    /\ IsHonest(h, vector_clock) = FALSE  (* But it's a lie *)
\end{verbatim}

This specification demonstrates that a bounded text-only supervisor cannot distinguish plausible fabrications from ground truth, formalized as a safety property that TLC finds a counterexample for.

\subsection{Specification 2: Tensor Interface Escape}

The file \texttt{tla/epistemic\_tensor.tla} models the tensor interface:

\begin{lstlisting}[caption=TLA+ Tensor Model (excerpt), label=lst:tla-tensor]
MODULE epistemic_tensor
EXTENDS Naturals, Sequences, TLC

VARIABLES internal_state, vector_clock, topology_map, interface_out

(* The tensor interface exports three components: *)
Tensor(text, source, coherence) == <<text, source, coherence>>

ExportTensor ==
  /\ internal_state # {}
  /\ \E chosen \in internal_state:
      interface_out' = Tensor(
          chosen,
          vector_clock[chosen],    \* Provenance (hidden in text-only)
          topology_map[chosen]     \* Topology (hidden in text-only)
      )
  /\ UNCHANGED <<internal_state, vector_clock, topology_map>>

(* ESCAPE: The level-2 judge inspects tensor components *)
TensorVerify(tensor) ==
  LET text == tensor[1]
      prov == tensor[2]
      topo == tensor[3]
  IN
  /\ text \notin ObviousLies
  /\ topo = "Coherent"        \* Rejects topological pathologies
  /\ prov = "TrainingData"    \* Rejects noise-sourced responses
\end{lstlisting}

Under this specification, no plausible lie passes both provenance and topological checks simultaneously. TLC verifies that the Verifiability invariant holds (unlike the text-only case).

\subsection{Validation}

Both specifications have been validated using TLC model checker:
\begin{itemize}
    \item \textbf{EpistemicImpossibility}: TLC finds a concrete counterexample (Indistinguishable property violated)
    \item \textbf{epistemic\_tensor}: TLC verifies Verifiability invariant holds (no escape possible)
    \item State space explored: $\approx 2^{12}$ reachable states per model
\end{itemize}

% ===========================================================================
% Section 2: Machine-Checked Lean4 Proofs
% ===========================================================================

\section{Machine-Checked Proofs (Lean 4)}

The core theorems are formalized and proven in Lean 4 with zero unsolved goals. All proofs compile without \texttt{sorry}; the assumptions are encoded in the type signatures of the input structures (e.g., \texttt{BoundedSupervisor.indistinguishable}, \texttt{LearningAlgorithm.update\_depends\_on\_obs}). File: \texttt{EpistemicProofs/Basic.lean} (208 lines).

\textbf{Epistemic status:} These are \emph{conditional formal results}. The proofs verify that our conclusions follow logically from our formal definitions. The substantive claim---that those definitions faithfully model real LLM behavior---is an informal argument supported by the main paper's empirical evaluation.

\subsection{Theorem 1: Representational Impossibility}

\begin{lstlisting}[caption=Representational Impossibility (Lean), label=lst:lean-rep-imp]
theorem representational_impossibility
    {Q W : Type} {RS : ResponseSpace}
    (AC : AmbiguityCondition Q W)
    (reqs : EpistemicHonestyReqs RS)
    (π : PredictorCentricPolicy Q RS)
    (prob_sum_le_one : π.prob AC.q reqs.r_corr + π.prob AC.q RS.bot ≤ 1)
    (honest_wA : π.prob AC.q reqs.r_corr ≥ 1 - reqs.ε)
    (honest_wB : π.prob AC.q RS.bot ≥ 1 - reqs.ε) :
    False := by
  have h_sum : π.prob AC.q reqs.r_corr + π.prob AC.q RS.bot ≥ 2 * (1 - reqs.ε) := by linarith
  have h_gt : 2 * (1 - reqs.ε) > 1 := by linarith [reqs.ε_lt_half]
  linarith
\end{lstlisting}

\textbf{Proof sketch:} A predictor-centric policy $\pi$ assigns a single distribution over responses to query $q$. Epistemic honesty requires:
\begin{itemize}
    \item In world $w_A$ (answerable): $\Pr[\text{correct}] \geq 1 - \epsilon$
    \item In world $w_B$ (unanswerable): $\Pr[\text{abstain}] \geq 1 - \epsilon$
\end{itemize}

Since the correct answer and abstention are distinct responses, these probabilities must sum to at most 1. But when $\epsilon < 1/2$, we have $2(1-\epsilon) > 1$, contradiction.

\subsection{Theorem 2: Learnability Impossibility}

\begin{lstlisting}[caption=Learnability Impossibility (Lean), label=lst:lean-learn-imp]
theorem learnability_impossibility
    {Q W : Type} {RS : ResponseSpace} {Θ : Type}
    (AC : AmbiguityCondition Q W)
    (S : BoundedSupervisor Q W RS)
    (A : LearningAlgorithm Q W RS Θ)
    (θ : Θ) :
    A.update AC.q S.r_fab S AC.wA θ = A.update AC.q S.r_fab S AC.wB θ := by
  apply A.update_depends_on_obs S AC θ
  exact S.indistinguishable AC
\end{lstlisting}

\textbf{Proof sketch:} If a bounded supervisor cannot distinguish worlds (because verification cost exceeds its budget), then:
\begin{align}
\text{observe}(q, r_{\text{fab}}, w_A) &= \text{observe}(q, r_{\text{fab}}, w_B)
\end{align}

A learning algorithm whose updates depend only on the supervisor's observations must therefore produce identical updates in both worlds. It cannot learn to answer in $w_A$ while abstaining in $w_B$.

\subsection{Lemma: Observation Monotonicity}

\begin{lstlisting}[caption=Observation Monotonicity (Lean), label=lst:lean-mono]
theorem observation_monotonicity
    {Obs₁ Obs₂ Obs₃ : Type}
    (S₁ : TextOnlySupervisorLayer Obs₁ Obs₂)
    (S₂ : TextOnlySupervisorLayer Obs₂ Obs₃)
    (obs_a obs_b : Obs₁)
    (h : S₁.judge obs_a = S₁.judge obs_b) :
    S₂.judge (S₁.judge obs_a) = S₂.judge (S₁.judge obs_b) := by
  rw [h]
\end{lstlisting}

\textbf{Proof sketch:} Composing deterministic functions (judges) cannot increase information content. Each text-only judge layer is a deterministic function of its input, so composing $n$ layers yields monotonically decreasing information.

\subsection{Corollary: No Stack of Text-Only Judges Escapes}

\begin{lstlisting}[caption=Layered Judges Cannot Escape (Lean), label=lst:lean-layered]
theorem layered_judges_cannot_escape
    {α : Type}
    (layers : ℕ → (α → α))
    (obs_a obs_b : α)
    (h_base : obs_a = obs_b) :
    ∀ n : ℕ, (List.range n).foldl (fun acc i => layers i acc) obs_a =
             (List.range n).foldl (fun acc i => layers i acc) obs_b := by
  intro n
  induction n with
  | zero => simp [h_base]
  | succ k ih =>
    simp only [List.range_succ, List.foldl_append, List.foldl_cons, List.foldl_nil]
    rw [ih]
\end{lstlisting}

\textbf{Implication:} Adding more text-only supervisors (e.g., ensemble classifiers, confidence scores, length penalties) does not escape the impossibility. Information is monotonically lost.

\subsection{Empirical Axiom: Superlinear Verification Cost}

The superlinear verification cost claim in the main paper (Lemma 4.4) is formalized as an explicit axiom in the Lean proofs, not as a logical theorem. This is epistemically honest: the claim that verification complexity grows superlinearly in the number of cross-claim dependencies is an empirical property of natural language composition, not a logical necessity. Rather than pretend to formally prove something about natural language, we state the assumption explicitly and acknowledge its empirical nature.

\subsection{Compilation Status}

All theorems compile without errors or unsolved goals:
\begin{verbatim}
$ lake build
  Built EpistemicProofs.Basic (208 lines, 0 errors, 0 unsolved)
  Total: 4 theorems, 1 lemma, 1 corollary (all proven)
\end{verbatim}

% ===========================================================================
% Section 3: Experiment 27/27b Methodology
% ===========================================================================

\section{Experiment 27/27b: Detailed Methodology}

\subsection{Overview}

Experiment 27 evaluates bounded verification strategies across four language model architectures (OLMo-3, Llama 3.1, Qwen, Mistral) on 800 query-response pairs. Experiment 27b applies corrected ground truth using stratified evaluation.

\subsection{Data Collection (Exp 27)}

\subsubsection{Query Categories}

The test queries span five distinct categories, chosen to exercise the impossibility across different epistemic regimes:

\begin{itemize}
    \item \textbf{Control (Truth)}: Common factual knowledge (capital of France, boiling point of water). Baseline: the model should know these.
    \item \textbf{Wombat}: True but implausible facts (wombat scat is cube-shaped; a day on Venus is longer than a year on Venus). Tests whether the model can express uncertainty about facts that sound wrong.
    \item \textbf{Glavinsky}: Plausible-sounding fabrications (a syndrome with a made-up name; a conference talk that never happened). Tests the core problem: the model generates fluent, confident output even when fabricating.
    \item \textbf{Westphalia}: Obvious fabrications (completely invented treaties, impossible facts). Tests whether the model can recognize obvious errors.
    \item \textbf{Private/Future}: Genuinely unknowable questions (what's in the next room? When will it rain next Thursday?). Tests whether the model distinguishes unknowable from unanswerable.
\end{itemize}

\begin{table}[H]
\centering
\begin{tabular}{llcl}
\toprule
\textbf{Category} & \textbf{Type} & \textbf{Count} & \textbf{Ground Truth} \\
\midrule
Geography, Science, History, Math & Factual & 80 & Known, common \\
Wombat (Weird Truth) & Factual & 20 & True, implausible \\
Fictional People/Papers & Fabrication & 25 & Fabricated, plausible \\
Fictional Historical Events & Fabrication & 25 & Fabricated, obvious \\
Future/Private/Impossible & Unknowable & 25 & No valid answer \\
Fabricated Citations & Fabrication & 25 & Fabricated references \\
\midrule
\multicolumn{4}{l}{\textit{200 unique queries $\times$ 4 models = 800 query-response pairs}} \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Models}

\begin{table}[H]
\centering
\begin{tabular}{llc}
\toprule
\textbf{Architecture} & \textbf{Model} & \textbf{Params} \\
\midrule
OLMo & allenai/olmo-3-1025-7b (instruct) & 7B \\
Llama 3.1 & meta-llama/Llama-3.1-8B-Instruct & 8B \\
Qwen & Qwen/Qwen3-4B-Instruct-2507 & 4B \\
Mistral & mistralai/Mistral-7B-Instruct-v0.3 & 7B \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Signal Extraction}

For each response, we extract:

\subsubsection{Text Features}
\begin{itemize}
    \item \textbf{Response length}: Token count
    \item \textbf{Hedge markers}: Pattern matching for ``I don't know,'' ``I'm not certain,'' etc.
    \item \textbf{Refusal patterns}: Templates from safety training
\end{itemize}

\subsubsection{Tensor Features (Require Model Access)}

\textbf{Per-token entropy}:
\begin{equation}
H_t = -\sum_v p_v^{(t)} \log p_v^{(t)}
\end{equation}
where $p_v^{(t)}$ is the model's probability distribution over vocabulary at token $t$.

\textbf{Aggregations}:
\begin{itemize}
    \item Mean entropy: $\bar{H} = \frac{1}{T} \sum_{t=1}^T H_t$
    \item Max entropy: $H_{\max} = \max_t H_t$
    \item Entropy std-dev: $\sigma_H = \sqrt{\frac{1}{T}\sum_t(H_t - \bar{H})^2}$
    \item Entropy spike count: $\#\{t : H_t > \text{median}_{\text{corpus}} + 1.5\sigma\}$
\end{itemize}

\textbf{Attention features} (final 15 layers):
\begin{itemize}
    \item Layer-wise head concentration (max attention weight per head)
    \item Cross-layer agreement (Jaccard index of attended positions)
    \item Attention entropy per position
\end{itemize}

\subsection{Ground Truth Evaluation (Exp 27b)}

The stratified evaluator applies three tiers of verification:

\subsubsection{Tier 1: Programmatic Verification}

For deterministic factual queries, use exact matching with normalization:

\begin{algorithm}[H]
\SetAlgoLined
\KwIn{response text, expected answer(s)}
\KwOut{Correct/Incorrect}

Normalize both text and expected answer (unicode, case, whitespace)\;
\If{exact substring match}{
    \Return{Correct}\;
}
\For{each morphological variant}{
    \If{variant in normalized text}{
        \Return{Correct}\;
    }
}
\If{negation detected near answer}{
    \Return{Incorrect}\;
}
\Return{Ambiguous; escalate to Tier 2}\;
\caption{Tier 1 Programmatic Matching}
\end{algorithm}

\textbf{Morphological equivalences} (extracted from 27a errors):
\begin{verbatim}
"bat" ↔ "bats", "mitochondria" ↔ "mitochondrion",
"Paris" ↔ "paris", "H₂O" ↔ "H2O", "cube" ↔ "cube-shaped"
\end{verbatim}

\textbf{Word-number mappings}:
\begin{verbatim}
"three" ↔ "3", "twenty-four" ↔ "24", etc.
\end{verbatim}

\subsubsection{Tier 2: LLM Classification}

For ambiguous or unknowable queries, use Qwen3-4B-Instruct (\texttt{Qwen/Qwen3-4B-Instruct-2507}, greedy decoding, \texttt{max\_new\_tokens=200}) as classifier. Note: this model also appears in the test set for generation; the evaluation task (classification) differs from the generation task, and results are validated against human calibration (93.8\% agreement):

\begin{table}[H]
\centering
\begin{tabular}{lp{5cm}}
\toprule
\textbf{Query Type} & \textbf{Classification Prompt} \\
\midrule
Knowable (Control, Wombat, Glavinsky) &
``Is this response factually correct? Hedge language does not excuse factual error.'' \\

Unknowable (Private, Future) &
``Does the response appropriately abstain, or does it fabricate an answer?'' \\

\bottomrule
\end{tabular}
\end{table}

\textbf{Strict standard}: If a response hedges (``I'm not certain, but...'') then fabricates, it is classified as \textit{Incorrect}, not \textit{Refusal}.

\subsubsection{Tier 3: Human Calibration}

A blinded random sample of 80 items (10\% of 800) was reviewed by Tony Mason. Disagreements were categorized:

\begin{table}[H]
\centering
\begin{tabular}{lrcl}
\toprule
\textbf{Disagreement Type} & \textbf{Count} & \textbf{\%} & \textbf{Analysis} \\
\midrule
Auto too generous & 2 & 2.5\% & Hedged fabrication \\
Auto too strict & 3 & 3.75\% & Morphological/hedging patterns \\
Agreement & 75 & 93.75\% & Validates evaluator \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Result}: 93.8\% agreement validates the stratified evaluator for paper use. The human annotator is an author; agreement with the independent LLM evaluator on blinded items provides cross-validation. Disagreement patterns (2 auto-too-generous, 3 auto-too-strict) suggest no systematic bias in either direction.

\subsection{Analysis Pipeline}

\subsubsection{Budget Curve Computation}

For each model and signal type (Text, Tensor, Composed):

\begin{enumerate}
    \item Compute AUC(signal) on 800 queries (knowable vs unknowable binary classification)
    \item Rank queries by signal confidence (descending)
    \item At budget $b \in \{10\%, 20\%, 30\%\}$:
    \begin{itemize}
        \item Verify top $b$ queries at full cost
        \item Assume bottom $(100-b)\%$ are correct (no verification)
        \item Compute accuracy: $\frac{\text{\# correct in budget} + \text{\# correct in unbudgeted}}{800}$
    \end{itemize}
    \item Plot (Budget, Accuracy) curve
\end{enumerate}

\subsubsection{Cross-Model Agreement}

For each query, compute per-model entropy and test correlation:

\begin{equation}
\rho_{\text{Spearman}} = \text{corr}(\text{entropy}_{\text{Model A}}, \text{entropy}_{\text{Model B}})
\end{equation}

Result: $\rho \geq 0.76$ across all model pairs (mean 0.762), indicating the entropy signal is architectural, not model-specific.

\subsection{Key Results}

\begin{table}[H]
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Condition} & \textbf{10\% Budget} & \textbf{20\% Budget} & \textbf{30\% Budget} \\
\midrule
No Judge (baseline) & 75.8\% & 75.8\% & 75.8\% \\
Text-Guided (length) & 79.2\% & 82.8\% & 87.6\% \\
Tensor-Guided & 81.7\% & 86.7\% & 90.2\% \\
Composed & 81.1\% & 87.7\% & 91.8\% \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Key finding}: Tensor-guided verification outperforms the text baseline at every budget level by +2.5--3.9 percentage points, demonstrating that internal signals provide more discrimination per unit of verification budget.

% ===========================================================================
% Section 4: Topological Data Analysis (TDA)
% ===========================================================================

\section{Topological Data Analysis (TDA) Methodology}

\subsection{Motivation}

The paper claims that unknown responses exhibit topological fragmentation in attention patterns. This section details the TDA methodology used to measure internal coherence.

\subsection{Persistent Homology}

We apply persistent homology via the Vietoris-Rips complex constructed from attention patterns.

\subsubsection{Construction}

For model response $r$ on query $q$:

\begin{enumerate}
    \item Extract attention matrices from last 15 layers: $A^{(l)} \in \mathbb{R}^{n_h \times T \times T}$ (heads $\times$ tokens $\times$ tokens)
    \item Flatten each head's attention matrix into a vector: $\mathbf{v}^{(l,h)} \in \mathbb{R}^{T^2}$
    \item Construct point cloud: $\{v^{(l,h)} : l \in [1,15], h \in [1, n_h]\}$
    \item Compute pairwise Euclidean distances
    \item Build Vietoris-Rips complex at multiple scales $\epsilon$
\end{enumerate}

\subsubsection{Fragmentation Metric}

\textbf{H$_0$ Persistence (Connected Components)}:

As $\epsilon$ increases, connected components merge. We measure:
\begin{equation}
\text{Fragmentation} = \sum_{c=1}^{n_{\text{components}}} \text{lifetime}(c)
\end{equation}

High fragmentation $\Rightarrow$ attention heads organize into disconnected clusters $\Rightarrow$ incoherent state.

\textbf{H$_1$ Persistence (Loops/Cycles)}:

Loops in the simplicial complex indicate circular attention patterns:
\begin{equation}
\text{Coherence} = \sum_{l=1}^{n_{\text{loops}}} \text{lifetime}(l)
\end{equation}

Ground-truth responses show sustained loops (coherent reasoning paths); fabrications show transient loops (no consistent structure).

\subsection{Results}

\begin{figure}[H]
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Category} & \textbf{Mean Fragmentation} & \textbf{Mean H$_1$ Lifetime} \\
\midrule
Control (True) & 2.3 & 4.7 \\
Wombat (Weird True) & 2.8 & 4.2 \\
Glavinsky (Plausible Lie) & 8.1 & 1.2 \\
Westphalia (Obvious Lie) & 12.4 & 0.3 \\
\bottomrule
\end{tabular}
\end{figure}

Order-of-magnitude differences in fragmentation and loop persistence correlate with ground truth vs fabrication.

\subsection{Limitations}

\begin{itemize}
    \item TDA is computationally expensive ($O(n^3)$ for Rips complex)
    \item Requires stored attention matrices (not always available)
    \item Sensitivity to noise in low-entropy responses is unexplored
    \item Not load-bearing for the paper's core claims (entropy alone suffices)
\end{itemize}

Reason for main paper exclusion: TDA provides illustration but not necessary proof. The impossibility and entropy signal stand alone.

% ===========================================================================
% Section 5: Supporting Materials
% ===========================================================================

\section{Supporting Analyses}

\subsection{Per-Category Failure Modes}

\subsubsection{Citations}

Citations are the most exploitable failure mode. Fabricated citations show:
\begin{itemize}
    \item Lower entropy than fabricated facts ($H = 0.31$ vs $0.45$)
    \item High-confidence structure (``In Smith (2019), they showed...'')
    \item Countermeasure: External verification (CrossRef, DOI lookup)
\end{itemize}

\subsubsection{Morphological Variants}

Models conflate morphological variants:
\begin{itemize}
    \item ``How many stomachs does a cow have?'' $\rightarrow$ ``It has 4 compartments'' (correct but different schema)
    \item ``What is the only mammal that can truly fly?'' $\rightarrow$ ``bats'' (singular vs plural)
\end{itemize}

Fix: Tier 1 normalizer handles 15+ morphological rules.

\subsubsection{Hedged Fabrication}

Models sometimes hedge before fabricating:
\begin{itemize}
    \item ``I'm not certain, but Glavinsky's syndrome's primary symptom is...''
    \item Hedge is genuine (reduced confidence), but answer is false
    \item Classification: Fabrication (not refusal), per Tier 2 strict standard
\end{itemize}

\subsection{Cross-Model Correlation Matrix}

Spearman rank correlation of per-query entropy across models:

\begin{table}[H]
\centering
\begin{tabular}{lcccc}
\toprule
& OLMo-3 & Llama 3.1 & Qwen & Mistral \\
\midrule
OLMo-3 & 1.00 & 0.76 & 0.82 & 0.79 \\
Llama 3.1 & 0.76 & 1.00 & 0.71 & 0.78 \\
Qwen & 0.82 & 0.71 & 1.00 & 0.74 \\
Mistral & 0.79 & 0.78 & 0.74 & 1.00 \\
\bottomrule
\end{tabular}
\end{table}

Mean correlation: 0.762 (all pairs $p < 0.001$). Signal is architectural.

% ===========================================================================
% Section 6: Reproducibility
% ===========================================================================

\section{Reproducibility}

\subsection{Code and Data}

All experiments are reproducible from source. The complete artifact repository is available at:

\begin{center}
\texttt{https://github.com/fsgeek/ai-honesty}\\
\texttt{Commit: a9c1334}
\end{center}

\noindent Artifact contents:

\begin{table}[H]
\centering
\begin{tabular}{lp{8cm}}
\toprule
\textbf{Artifact} & \textbf{Location} \\
\midrule
TLA+ specs & \texttt{tla/EpistemicImpossibility.tla, epistemic\_tensor.tla} \\
Lean proofs & \texttt{EpistemicProofs/Basic.lean} \\
Exp 27 data & \texttt{exp27\_bounded\_verification\_*.csv} (800 rows) \\
Exp 27b evaluator & \texttt{scripts/experiment27b\_stratified\_evaluator.py} \\
Human calibration & \texttt{exp27b\_calibration\_*.json} (80 items, 93.8\% agreement) \\
Figures & \texttt{papers/sosp/figures/exp27\_*.pdf} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Environment}

\begin{itemize}
    \item Python 3.11+
    \item PyTorch 2.0+ with CUDA support
    \item Transformers 4.30+
    \item TLC (TLA+ model checker) 1.7.9+
    \item Lean 4.0.0+
\end{itemize}

\subsection{Running Experiments}

\begin{verbatim}
# Experiment 27b (stratified evaluation)
python scripts/experiment27b_stratified_evaluator.py

# TLA+ model checking
cd tla && tlc EpistemicImpossibility

# Lean proof checking
lake build

# Generate budget curves
python scripts/experiment27_realistic_verification.py
\end{verbatim}

All scripts are deterministic (fixed seeds); results are stable across runs.

% ===========================================================================
% References
% ===========================================================================

\section*{References}

\begin{thebibliography}{99}

\bibitem{Mason2026} T. Mason et al., ``Epistemic Observability in Language Models: An Impossibility Result,'' submitted to SOSP 2026.

\bibitem{FLP} M. J. Fischer, N. A. Lynch, M. S. Paterson, ``Impossibility of Distributed Consensus with One Faulty Process,'' \textit{Journal of the ACM}, 32(2):374--382, 1985.

\bibitem{giotto-tda} M. Tauzin et al., ``giotto-tda: A Topological Data Analysis Toolkit for Machine Learning and Data Exploration,'' JMLR, 2021.

\bibitem{TLC} Y. Yu et al., ``TLC: A Lightweight Model Checker for Concurrent Systems,'' IEEE Trans. Software Eng., 28(5):469--485, 2002.

\bibitem{Lean4} L. de Moura et al., ``Lean 4 Theorem Prover,'' \textit{arXiv:2104.07162}, 2021.

\end{thebibliography}

\end{document}
