%% ==========================================================================
%% SECTION 1: INTRODUCTION
%% ==========================================================================
\balance
\section{Introduction}\label{sec:intro}

Systems incorporating language model components face a verification problem.
The components generate text by sampling from learned probability distributions,
and some fraction of that text is fabricated: fluent, confident, and wrong.
LLM-based systems today are rife with ``hallucinations''.
We use the term \emph{fabrication} throughout this paper to emphasize that
the system's default behavior under coherence optimization is reasonable and
not a malfunction to be patched.
\citeauthor{xu2024hallucination}~\cite{xu2024hallucination} prove that fabrication is
inevitable for LLMs over the class of computable functions.

% This paper provides the formal framework and empirical cost surface for making that decision.

% We use the term \emph{fabrication} throughout this paper to emphasize that
% the system's default behavior under coherence optimization is reasonable and
% not a malfunction to be patched. Language models optimized for
% coherence produce coherent output. The output reflects the training data: grounded
% training data tends to track truth, while ungrounded training data tends to produce untruth.
% Our result is orthogonal. They prove a computational limit on what the model can produce;
% we prove an observational limit on what a supervisor can verify.
% [moved to telemetry paragraph below]

%Verification has a cost. 
Consequently, every system that
deploys a language model component must allocate a budget: how much of the output to verify,
which outputs to prioritize, and what signals to use for triage.
Verification cost varies by mechanism and content type: checking a citation
against a database costs latency; running a second model as a judge costs
compute; assessing whether a summary faithfully represents its source is
more expensive still.

The fundamental source of these costs is the interface through which language models are observed. 
Today's systems expose only the generated text while hiding the model's internal computation,
which contains signals that may distinguish grounded generation from fabrication. 
Because these signals are discarded before reaching the supervisor, they cannot be recovered from the output alone. 
In an accompanying technical report~\cite{mason2026epistemic}, we show that, under bounded supervision, 
grounded and fabricated internal states can be observationally equivalent through the text channel alone, 
independent of model scale or training procedure.

This observability gap has direct consequences for verification cost:
a supervisor limited to text must expend its verification budget reconstructing information 
that was available during generation but hidden by the interface, instead of focusing verification 
effort on the small fraction of outputs that genuinely require additional scrutiny. 
This paper therefore asks the systems question: if language models exposed richer observational interfaces, 
how would that change the cost, effectiveness, and return on investment of verification?

Our key insight is that these hidden signals need not remain hidden. 
During inference, language models naturally produce auxiliary information that reflects their internal computation. 
We refer to these signals as \emph{telemetry}: byproducts of inference that the model does not explicitly choose to communicate. 
This distinguishes telemetry from the model's \emph{testimony}---the generated text itself. 
Whereas testimony is optimized to satisfy the prompting objective and can therefore be misleading, 
telemetry arises as a consequence of computation and is correspondingly harder to manipulate.

We introduce \emph{epistemic observability}: an interface that exposes this
telemetry alongside the standard text output.
Unlike interpretability, which seeks to explain why a model activates particular neurons, 
or explainability, which generates post-hoc rationales for its outputs, 
epistemic observability surfaces signals produced as a byproduct of inference that help 
estimate whether the generated content is grounded in the model's knowledge or is likely to be fabricated.

To support a range of applications, we propose a multi-tier
epistemic observability interface, where each successive tier exposes richer
telemetry about the generation process at greater computational cost.
A creative brainstorming assistant may require only Tier~0 observability,
whereas a medical decision-support system may justify the cost of higher
tiers; the hierarchy makes explicit what additional epistemic information
each level of investment provides.

As a proof of concept, we design a tensor-generation interface that exposes
lightweight epistemic telemetry---per-token entropy (Tier~1) and attention
summaries (a low-cost Tier~2 signal)---and demonstrate that tensor-guided judges consistently outperform text-only baselines across all verification budgets, 
improving accuracy by 3.2--4.7 percentage points on average across four model architectures. 

% \fakepara{Empirical cost surface of verification.} We map the cost
% surface across text-only and tensor-augmented verification strategies.
% Under length-controlled evaluation, no text-channel feature exceeds pooled
% AUC 0.70; per-token entropy is the only signal that generalizes across
% architectures. Tensor-guided triage outperforms text baselines at every
% budget level (\autoref{sec:eval}).
